Challenges to transboundary fisheries management in North America under climate change

### 2020-01-24 Manuscript_Draft.Rmd
Started corrections on manuscript

---
title: "Challenges to transboundary fisheries management in North America under climate change"
author: "Juliano Palacios-Abrantes^1^ U. Rashid Sumaila^1^ William W. L. Cheung^1^"
subtitle: "[THIRD DRAFT]"
date: "^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada"
output:
  word_document: default
  pdf_document: default
bibliography: Oceans_Canada_Biblio.bib
editor_options: 
  chunk_output_type: console
csl: harvard.csl
---

@@

# End Century
End_In <- paste(2080);End_In # 2090
End_End <- paste(2099);End_End
```

@@

Climate change is shifting the distribution of fish stocks that straddle between Exclusive Economic Zones (EEZ), challenging transboundary fisheries management. In this paper, we examine the projected sharing of jointly managed transboundary fish stocks between the EEZs of Canada and the United States. We hypothesize that ocean warming will alter the sharing of fish stocks between the two countries, and that such changes will intensify under a high carbon emission scenario impacting the stock-share ratio of different stocks. Firstly, we projected changes in potential catch of 33 transboundary stocks of North America in the 21^st^ century based on multiple Earth system models’ simulations and species distribution models, under two climate change scenarios. We then look at the specific cases of the International Pacific Halibut Commission that manages pacific halibut (*Hippoglossus stenolepis*) and a resource sharing agreement in the Gulf of Maine for cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*) and yellowtail flounder (*Limanda ferruginea*). Results show that, even under a low emission scenario, most transboundary fish stocks sharing ratios, i.e., the proportion of the total catch of a fish stock taken by a given country, will change by 2050 relative to present in the direction as expected from the effects of ocean warming. The overall reduction in catch potential, in addition to the changes in stock-share will further exacerbate trade-offs between changes in species catch potential. Such trade-offs in the Atlantic and Pacific regions will be amplified if a high emission scenario is followed, relative to a low carbon emission scenario. Our paper highlights the challenges that transboundary fisheries management will face as species shift their current distribution under a changing climate.[^Acknowledgements]

@@

<!-- ##### First paragraph of paper summary -->
Transboundary fisheries are defined as fish stocks that move freely between (*i*) Countries economic exclusive zones (EEZs) and (*ii*) between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. The management of transboundary stocks can be analized using game theory, as often, success depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Therefore, in 1985, the United Nations incentivized actions to cooperate on the management of transboundary fisheries through the Law of the Seas (UNCLOS) [@UnitedNations:1986tl]. Today, an estimated 347 [@Teh:2015gd] to 1500 [@Caddy:1997ue] fish stocks cross national borders, some of them jointly managed by two or more countries. These stocks are responsible for almost 50% of these countries total fish catches [@Teh:2015gd]. Transboundary stocks such as salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) are of utmost importance for Canada and the United States (US). Management of these stocks is extremely hard due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf].

@@

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd]. To cope with these changes in ocean biophysic properties, marine species, including transboundary fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], transboundary fisheries' delicate governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tropical tunas, have significantly increased in subtropical regions of the Atlantic and western Pacific Oceans [@MonllorHurtado:2017cm]. Multiple transboundary species in North America have been observed to shift in distribution following changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the 21^st^ century [@Cheung:2018hz] at a rate of tens to hundreds of km and/or tens of meters per decade towards the poles and/or deeper waters, respectively, although there will be regional variations [@Morley:2018fn]. As a result, some countries or management jurisdictions may see more transboundary stocks and their catches shifting into their waters while others will stand to lose [@Miller:2013iv; @Pinsky:2018cb]. Nevertheless, management rules for shared stocks (e.g. quota or spatial delimitation) are often determined based on current and historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp]. In addition, many international treaties do not directly consider climate change in their management scheme, despite evidence of target species shifting their distribution.

@@

Transboundary fisheries between Canada and the US offer a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of these fisheries. Canada and the US have a long history of fisheries cooperation [@Song:2017va]. They participate in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@Merten:2015tu]. Cooperation schemes vary by fisheries stocks, regulatory areas, legal terms of agreement, and management measures, among others (Fig. 1). In the Atlantic coast, Canada and the US are part of the Northwest Atlantic Fisheries Organization (NAFO) that oversees the management of 19 stocks of 11 transboundary species (although around 41 taxa are fished within Canadian and US waters) from southern Gulf of Maine to Greenland[^NAFO]. Within NAFO's region 5Z, Canada and the US have a "Resource Sharing Understanding" (hereafter referred to "GoM agreement") to set catch-limits of Atlantic cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Soboil:2006gw; @CIA:2017uta]. In the Pacific coast, Pacific halibut (*Hippoglossus stenolepis*) is managed under the International Pacific Halibut Commission (IPHC) [@IPHCSecreatriat:2017uz]. All five salmon species (chinook, *Oncorhynchus tshawytscha*; chum, *O. keta*; coho, *O. kisutch*; pink, *O. gorbuscha*; and sockeye *O. nerka*), are managed by the Pacific Salmon Commission [@PSC:2017tb]. Finally, under the Pacific Whiting Treaty, Canada and the United states jointly manage the coastal stock of this Pacific hake species (*Merluccius productus*) [@Merten:2015tu].

@@

@@

![Fig 1. Map of North America with the regulatory areas of the IPHC and the NAFO sub-region related to the Gulf of Main agreement](./Results/Fig1.png)

It is expected that climate induced shifts in stock distribution affect the rules in place that keeps these treaties alive. Therefore, the main objective of this paper is to assess the level of exposure of international fisheries management treaties between Canada and the US to climate change impacts through shifts in stock distributions. Specifically, we rely on a species distribution model and scenario planning to project the changes in the distribution of 33 fish stocks jointly managed by Canada and the US. Moreover, we use two specific treaties as case studies (IPHC and the GoM agreement) to explore the potential impacts that climate change will have on selected transboundary fisheries management. Finally, we explore similar situations around the world and identify opportunities to improve the adaptability of transboundary fisheries management to climate change in North America. Despite an overall expectation of species following a poleward shift, important geographic features (e.g. Gulf of Alaska representing a latitudinal block), geo-political (e.g. the localization of Alaska in reference to Canada and the contiguous states), and management rules (e.g. quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision makers on the paths to follow under a changing climate.

The current study focuses on transboundary fisheries jointly managed by Canada and the US for both the Atlantic and the Pacific coasts. The analysis was carried out within the 200 nautical miles of both nation's EEZ from the Gulf of Mexico to The Labrador Sea in the Atlantic, and from California to Chukchi Sea, in the Pacific coast (Fig. 1). Subsets of the total area were carried for both the IPHC and the Gulf of Maine agreement. For the IPHC, we used the most updated spatial regulatory data provided by the Commission [@IPHCSecreatriat:2017uz]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoM agreement we used NAFO's divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Fig. 1).

@@

@@

path_world <- "~/Documents/UBC/Oceans_Project/Manuscript/Data/Spatial_Data/TM_WORLD_BORDERS"
file_name <- "TM_WORLD_BORDERS"

@@

path_eez <- ("~/Documents/UBC/Oceans_Project/Manuscript/Data/Spatial_Data/World_EEZ_v10_2018")
#The File
fnam_eez <- "eez_v10.shp"

@@

# Now we load IPHC shapefiles ####
path_IPHC <- ("~/Documents/UBC/Oceans_Project/Distribution/Data/IPHC/IPHC_RegulatoryAreas/")

@@

path_NAFO <- ("~/Documents/UBC/Oceans_Project/Distribution/Data/Spatial/NAFO/")
#The File
fnam_NAFO <- "Divisions"

@@

The species analyzed (**Table A2.1**) were selected based on five treaties in both the Atlantic and Pacific coast of North America (Table 1). All species under the IPHC, PSC, WT, and the GoM agreement were analyzed as these treaties are management agreements between Canada and the US. However, in the case of NAFO, their management covers over 41 taxa and its governance extends to international waters and the EEZ of Greenland. Therefore, a subset of 26 species reported was made based on those fished in NAFO regions that fall within the EEZs of Canada and the US (NAFO subareas 2's, 3's with the exception of 3M and 3N and limited to the EEZ, 4's,5's and 6 A,B, and C.) [@NAFO]. Data related to fisheries was gathered from the *Sea Around Us* from 1951 to 2014 [@Zeller2016].

@@

data.table::data.table(Treaty =
                         c("International Pacific Halibut Coomission (IPHC)",
                           "Pacific Salmon Coomission (PSC)",
                           "Pacific Whiting Treaty (PWT)",
                           "Northwest Atlantic Fisheries Organization (NAFO)",
                           "Gulf of Main's Agreement (GoM Agreement)"),
                       Species =
                         c(1,
                           5,
                           1,
                           26,
                           "3*"),
                       Coast =
                         c("Pacific",
                           "Pacific",
                           "Pacific",
                           "Atlantic",
                           "Atlantic")
) %>% 
  kable("latex",
        caption = "Treaties and number of species included in the current analysis."
        ) %>% 
  add_footnote(fn)
```

@@

where $X_{t}$ is the future agregated MCP for two periods, $t_1$ = mid 21^st^ century (2045-2055); and $t_3$ = end of 21^st^ century (2090-2099), and $t_0$ is the present aggregated MCP ($\mu$ 2005-2014).

@@

In addition, we borrowed the concept of “threat point” from game theory (see Sumaila et al., this Volume) and estimate the change in the $\hat{MCP}$ that each country would have for each species (hereafter referred as stock-share ratio), for both the IPHC and the GoM agreement. For this, we first modify equation 1, to estimate the aggregated yearly mean MCP of each species per region. We then average the results by the same previouslly metioned periods (present, mid and end of the 21^st^ century). Next, for each species we estimated the stock-share ratio ($\alpha$) that each region had during each time period:

@@

# List of DBEM INDEX within the EEZ
#Data provided by Vicki Lam

EEZIDs_List <- read_excel("~/Documents/UBC/Oceans_Project/Distribution/By Cell/Cell_Data/Updated_EEZList_17June2016.xlsx")

EEZ_CellID <- read_excel("~/Documents/UBC/Oceans_Project/Distribution/By Cell/Cell_Data/EEZ_CellID.xlsx")

# From (/Users/jpalacios/Documents/UBC/Oceans_Project/Distribution/By Cell/By_Cell_Analysis.Rmd)
Raw_DBEM_Results <- data.table::fread("~/Documents/UBC/Oceans_Project/Manuscript/Data/Raw_Data.csv") %>% 
  filter(Species != "Anoplopoma fimbria") #Remove

@@

### North American transboundary fisheries are shifting poleward

@@

#### New Future Method to account for mean +- sd of models #
MCP_Data <- DBEM_Corrected %>% 
  mutate(RCP = ifelse(Model %in% RCP,"Low_Emission","High_Emission")) %>% 
  group_by(INDEX,
           Model,
           RCP) %>% 
  summarise_at(vars(Beg_In:End_End), sum,na.rm=T) %>% # Sum the MCP of all species in each index by model
  # select(INDEX,Model,Beg_In:Beg_End) %>% 
  gather("Year","MCP",Beg_In:End_End) %>% 
  mutate(  # Indicate the three period times
    Period = ifelse(Year >= Beg_In & Year <= Beg_End,"Begining",
                    ifelse(Year >= Mid_In & Year <= Mid_End,"Mid_Century",
                           ifelse(Year >= End_In & Year <= End_End,"The_End_Century",
                                  "Other_Years"))
    )
  ) %>% 
  group_by(INDEX,Model,RCP,Period) %>% 
  summarise(Mean_T_MCP =mean(MCP,na.rm=T)) %>% #Average pero Time period
  filter(Period != "Other_Years") %>% 
  spread(Period,Mean_T_MCP) %>% #Spread to estimate MCP_Change
  mutate_at(vars(Mid_Century, # Devide future values by present
                 The_End_Century),
            funs(ifelse(Begining > 0, ./Begining,.)) #if no species today value is the future
  ) %>% 
  mutate_at(vars(Mid_Century, # Estimate percentage change
                 The_End_Century),
            funs(ifelse(Begining == 0 & . == 0, 0, # If for no change than 0
                        ifelse(Begining == 0 & . > 0, 100, # If changes from 0 to anything then 100
                               round(-(1-.)*100) # Estimate percentage change
                        )
            )
            )
  ) %>% 
  gather("Period","MCP_Change",Mid_Century,The_End_Century) %>% 
  mutate(Direction = ifelse(MCP_Change >= 0, "Positive","Negative")) %>% 
  group_by(INDEX,Period,RCP) %>% # Average results per ESMs
  summarise_at(vars(MCP_Change),
               funs("Mean_ESM_MCP_Chng"=mean,
                    "sd_ESM_MCP_Chng"=sd,
                    "N_ESMs"=n(),
                    "Model_Agreement" = paste(length(unique(Direction)),collapse = ";")
               )
  ) %>%
  mutate(
    Model_Agreement = ifelse(Model_Agreement == 1, "Agree","Disagree")
  ) %>% 
  left_join(DBEM_Coor,
            by="INDEX") %>% 
  mutate_at(vars(Mean_ESM_MCP_Chng,
                 sd_ESM_MCP_Chng),
            funs(ifelse(. > 100, 100,.)
            )  # set everything over 100 to 100
  )

@@

![Fig 2. Percentage change of MCP of transboundary fisheries of North America for mid century (2046-2055) referent to 2005-2014 under a A) low emission scenario and; B) high emission scenario.](./Results/Fig2.png)

@@

In consequence, country's share-ratio of different species would change in the direction as expected from the effects of ocean warming, regardless of the RCP scenario. Overall, shifts will be amplified under the high emission scenario relative to the low greenhouse emission scenario pathway (Table 2). While in 2050, the average shift of species will be higher for East Canada and lower for East US under the high emission scenario, the stock-share ratio of some species is expected to increase more than 50% as in the case of Atlantic menhaden (*Brevoortia tyrannus*) in Eastern Canada, and decrease by more than 50% for species such as Cusk (*Brosme brosme*) in the Eastern US under a low emission scenario. By the end of the 21^st^ century, losses will intensify with the eastern US potentially losing on average 15.2% or 46% of stock-share, depending on the scenario.

@@

@@

>>>>

---
title: "Challenges to transboundary fisheries management in North America under climate change"
author: "Juliano Palacios-Abrantes^1^ U. Rashid Sumaila^1^ William W. L. Cheung^1^"
subtitle: "[Response to Reviewers]"
date: "^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada"
output:
  word_document: default
  pdf_document: default
bibliography: Oceans_Canada_Biblio.bib
editor_options: 
  chunk_output_type: console
csl: harvard.csl
---

@@

# End Century
End_In <- paste(2080);End_In # 2090
End_End <- paste(2099);End_End

Data_Path <- "~/Documents/UBC/OC_Special_Issue/Ecol/Data"

@@

Climate change is shifting the distribution of fish stocks that straddle between Exclusive Economic Zones (EEZ), challenging transboundary fisheries management. In this paper, we examine the projected sharing of jointly managed transboundary fish stocks between the EEZs of Canada and the United States. We hypothesize that ocean warming will alter the sharing of fish stocks between the two countries, and that such changes will intensify under a high carbon emission scenario impacting the stock-share ratio of different stocks. Firstly, we projected changes in potential catch of 33 transboundary stocks of North America in the 21^st^ century based on multiple Earth system models’ simulations and species distribution models, under two climate change scenarios. We then look at the specific cases of the International Pacific Halibut Commission that manages pacific halibut (*Hippoglossus stenolepis*) and a resource sharing **DV arrangement** in the Gulf of Maine for cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*) and yellowtail flounder (*Limanda ferruginea*). Results show that, even under a low emission scenario, most transboundary fish stocks sharing ratios, i.e., the proportion of the total catch of a fish stock taken by a given country, will change by 2050 relative to present in the direction as expected from the effects of ocean warming. The overall reduction in catch potential, in addition to the changes in stock-share will further exacerbate trade-offs between changes in species catch potential. Such trade-offs in the Atlantic and Pacific regions will be amplified if a high emission scenario is followed, relative to a low carbon emission scenario. Our paper highlights the challenges that transboundary fisheries management will face as species shift their current distribution under a changing climate.[^Acknowledgements]

@@

<!-- ##### First paragraph of paper summary -->
**RA In 1982, the United Nations Law of the Seas (UNCLOS) formalized the concept of economic exclusive zones (EEZs) creating what we know today as transboundary stocks. Transboundary stocks move freely between (*i*) Countries EEZs and (*ii*) between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. While not a law, UNCLOS incentivized actions to cooperate on the management of transboundary stocks [@UnitedNations:1986tl]. Today, game theory is one of the most common approaches used to analyze the management of transboundary stocks, as often, success depends on effective cooperation between parties** [@Munro:2002uf; @Sumaila:2013hv]. An estimated 347 [@Teh:2015gd] to 1500 [@Caddy:1997ue] fish stocks cross national borders, some of them jointly managed by two or more countries. These stocks are responsible for almost 50% of these countries total fish catches [@Teh:2015gd]. Transboundary stocks such as salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) are of utmost importance for Canada and the United States (US). Management of these stocks **RA can be convoluted** due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf].

@@

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd]. To cope with these changes in ocean biophysic properties, marine species, including transboundary fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], transboundary fisheries' delicate governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tropical tunas, have significantly increased in subtropical regions of the Atlantic and western Pacific Oceans [@MonllorHurtado:2017cm]. Multiple transboundary species in North America have been observed to shift in distribution following changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the 21^st^ century [@Cheung:2018hz] at a rate of tens to hundreds of km and/or tens of meters per decade towards the poles and/or deeper waters, respectively, although there will be regional variations [@Morley:2018fn]. As a result, some countries or management jurisdictions may see more transboundary **RA fisheries** and their catches shifting into their waters while others will stand to lose [@Miller:2013iv; @Pinsky:2018cb]. Nevertheless, management rules for shared stocks (e.g. quota or spatial delimitation) are often determined based on current and historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp]. In addition, many international treaties do not directly consider climate change in their management scheme, despite evidence of target species shifting their distribution.

@@

Transboundary **RB stocks** between Canada and the US offer a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of these fisheries. Canada and the US have a long history of fisheries cooperation [@Song:2017va]. They participate in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@Merten:2015tu]. Cooperation schemes vary by fisheries stocks, regulatory areas, legal terms of agreement, and management measures, among others. In the Atlantic coast, Canada and the US are part of the Northwest Atlantic Fisheries Organization (NAFO) that oversees the management of 19 stocks of 11 transboundary species (although around 41 taxa are fished within Canadian and US waters) from southern Gulf of Maine to Greenland [^NAFO]. In the Pacific coast, Pacific halibut (*Hippoglossus stenolepis*) is managed under the International Pacific Halibut Commission (IPHC) [@IPHCSecreatriat:2017uz]. All five salmon species (chinook, *Oncorhynchus tshawytscha*; chum, *O. keta*; coho, *O. kisutch*; pink, *O. gorbuscha*; and sockeye *O. nerka*), are managed by the Pacific Salmon Commission [@PSC:2017tb]. Finally, under the Pacific Whiting Treaty, Canada and the United states jointly manage the coastal stock of this Pacific hake species (*Merluccius productus*) [@Merten:2015tu].

@@

It is expected that climate induced shifts in stock distribution **RA will** affect the rules in place that **RA keep** these treaties alive. Therefore, the main objective of this paper is to assess the level of exposure of international fisheries management treaties between Canada and the US to climate change impacts through shifts in stock distributions. Specifically, we rely on a species distribution model and scenario planning to project the changes in the distribution of 33 fish stocks jointly managed by Canada and the US. Moreover, we use two specific treaties as case studies (IPHC and an **DV arrangement** for the Gulf of Maine) to explore the potential impacts that climate change will have on selected transboundary **RB stocks** management. Finally, we explore similar situations around the world and identify opportunities to improve the adaptability of transboundary **RB stocks** management to climate change in North America. Despite an overall expectation of species following a poleward shift, **RA important geographic constraints (e.g. Gulf of Alaska representing a latitudinal block) [@Kleisner:2016je], geo-political features** (e.g. the localization of Alaska in reference to Canada and the contiguous states), and management rules (e.g. quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision makers on the paths to follow under a changing climate.

The current study focuses on transboundary **RB stocks** jointly managed by Canada and the US for both the Atlantic and the Pacific coasts. The analysis was carried out within the 200 nautical miles of both nation's EEZ from the Gulf of Mexico to The Labrador Sea in the Atlantic, and from California to Chukchi Sea, in the Pacific coast. The species analyzed (**Table A2.1**) were selected based on five treaties in both the Atlantic and Pacific coast of North America (Table 1). All species under the IPHC, PSC, WT, and the Gulf of Maine **DV arrangement** were analyzed as these treaties are management agreements between Canada and the US. However, in the case of NAFO, their management covers over 41 taxa and its governance extends to international waters and the EEZ of Greenland. Therefore, a subset of 26 species reported was made based on those fished in NAFO regions that fall within the EEZs of Canada and the US (NAFO subareas 2's, 3's with the exception of 3M and 3N and limited to the EEZ, 4's,5's and 6 A,B, and C.) [@NAFO]. Data related to fisheries was gathered from the *Sea Around Us* from 1951 to 2014 [@Zeller2016].

We selected two case studies, the IPHC agreement and the Gulf of Maine arrangement (hereafter refered as GoM arrangement), in order to better discuss the implications that climate chage could have in the management of transboundary stocks. For the IPHC, we used the most updated spatial regulatory data provided by the Commission [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoM **DV arrangement** we used NAFO's divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Fig. 1).

@@

![Fig 1. Map of North America with the regulatory areas of the IPHC and the NAFO sub-region related to the Gulf of Main **DV arrangement**](./Figures/Fig1.png)

@@

# List of DBEM INDEX within the EEZ
#Data provided by Vicki Lam
EEZIDs_List <- read_excel(paste(Data_Path,"/Spatial/Updated_EEZList_17June2016.xlsx",sep=""))
EEZ_CellID <- read_excel(paste(Data_Path,"/Spatial/EEZ_CellID.xlsx",sep=""))
colnames(EEZ_CellID) <- c("EEZID","INDEX")

# DBEM Coordinate system
Coor <- fread(paste(Data_Path,"/Spatial/Lon_Lat_DBEM.txt",sep=""),header = FALSE)
colnames(Coor) <- c("INDEX","Longitude","Latitude")

@@

path_world <- paste(Data_Path,"/Spatial/TM_WORLD_BORDERS",sep="")
file_name <- "TM_WORLD_BORDERS"

@@

path_eez <- paste(Data_Path,"/Spatial/World_EEZ_v10_2018",sep="")
#The File
fnam_eez <- "eez_v10.shp"

@@

# Now we load IPHC shapefiles ####
path_IPHC <- paste(Data_Path,"/Spatial/IPHC_RegulatoryAreas/",sep="")

@@

path_NAFO <- paste(Data_Path,"/Spatial/NAFO",sep="")
#The File
fnam_NAFO <- "Divisions"

@@

##### The International Pacific Halibut Comission (IPHC)

The IPHC is a convention established in 1923 by Canada and the United States that oversees the management of Pacific halibut (*Hippoglossus stenolepis*), a flatfish that can grow as much as 500 pounds (2.4 m) which inhabits the continental shelf of the northern Pacific Ocean [@IPHC:2014tz]. The IPHC Convention Area covers the Pacific EEZ of Canada and the US from California to the Berring Sea [@IPHC:2019wt] and is divided in 12 regions (Fig 1). Most of the Pacific halibut population is estimated to be in region 3 (51.2%), followed by regions 2 and 4ACDE (23.1% and 20.4%, respectively), and lastly region 4B with only 5.2% of the stock distribution [@IPHC:2018vz]. In terms of management, the IPHC implements a total allowable catch (TAC) in addition to a series of regulations to control fishing effort. Every year, the IPHC carries an fishery-independent set-line survey from Oregon to the Berring sea to support the estimation of next fishing season's fishing limits [@IPHC:2018vz]. Thus, the TAC is allocated by region according to the population distribution with sector 3A historically having almost 1/3 of the TAC and almost half of the commercial catch. The TAC is divided between recreational, subsistence and commercial fishery, with a portion set aside for bycatch of other fisheries [@IPHC:2019wt]. The commercial fishing season starts in March ending around November whith restrictions allowing only set line gear with J-type hooks targeting individuals over 81.3 cm (32 inches) with head on [@IPHC:2019wt].

##### The Gulf of Maine Agreement

Since 1998 Canada and the US celebrate a "Resource Sharing Understanding" to manage three groundfish stocks within the Gulf of Maine (Georges Bank) [@TRAC:2016wu]. Under the GoM **DV arrangement**", both countries set catch-limits for George Bank's Atlantic cod (*Gadus morhua*), and Eastern George Bank's haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Pudden:2007bo; @CIA:2017uta]; **Koubrak and VanderZwaag, this Special Feature**. In terms of management, the quota allocation formula includes a combination of future stock's distribution patterns produced by yearly surveys and historical utilization patterns favoring Canda in terms of Cod (82%) and the US for Haddock (66%) and Y. flounder (66%), althou these patterns can change on yearly bases [@Pudden:2007bo; @TRAC:2016wu]. In addition, Canada and the US conduct jointly surveys and enforce activities of the maritime border. The task of the committee is just to inform Canada and the US who ultimatley take single management decissions. In terms of manangement, the US has a multi species harvest control with area and season cloasures, mesh sizes, effort control. US mobile gear vessels that use bottom otter trawl gear [@Soboil:2006gw]. In contrast, Canada has a more strict regulatory system implaced based on a quota system in addition to limited-entery licensig, fleet allocations, mesh sizes, fish size regulation, among other input controls [@Soboil:2006gw]. Canada inshore vessels fish Cod with longline and gillnet while haddock is mainly caught with bottom otter trawl gear [@Soboil:2006gw].

@@

data.table::data.table(Treaty =
                         c("International Pacific Halibut Coomission (IPHC)",
                           "Pacific Salmon Coomission (PSC)",
                           "Pacific Whiting Treaty (PWT)",
                           "Northwest Atlantic Fisheries Organization (NAFO)",
                           "Gulf of Main's Arrangement (GoM Arrangement)"),
                       Species =
                         c(1,
                           5,
                           1,
                           26,
                           "3*"),
                       Coast =
                         c("Pacific",
                           "Pacific",
                           "Pacific",
                           "Atlantic",
                           "Atlantic")
) %>% 
  kable("latex",
        caption = "Treaties and number of species included in the current analysis."
        ) %>% 
  add_footnote(fn)
```

@@

**RB where $X_{t}$ is the future averaged MCP for each of the two time periods analized in this study and $t_0$ is the present averaged MCP ($\mu$ `r Beg_In`-`r Beg_End`). This way, Equ. 2 shows the percentage change in MCP relative to today ($t_0$) by mid 21^st^ century when $X_T$ = $\mu$ `r Mid_In`-`r Mid_End`, and end of 21^st^ century when $X_T$ = $\mu$ `r End_In`-`r End_End`). The rationale between choosing these time periods was to provide a relative short-term projection (mid-century) that would be more policy-relevant but also show the long term trend.**

@@

In addition, we borrowed the concept of “threat point” from game theory **RA defined as the minimum payoff that a player is willing to receive in order to cooperate with other players** (see Sumaila et al., this Volume). Thus we estimate the change in the $\hat{MCP}$ **RA (threat point)** that each country **RA (players)** would have for each species (hereafter referred as stock-share ratio), for both the IPHC and the GoM **DV arrangement**. **RA We define stock-share ratio as the proportion of the stock's distribution within the study area that each country has**. For this, we first modify equation 1, to estimate the aggregated yearly mean MCP of each species per region. We then average the results by the same previouslly metioned periods (present, mid and end of the 21^st^ century). Next, for each species we estimated the stock-share ratio ($\alpha$) that each region had during each time period:

@@

# List of DBEM INDEX within the EEZ
#Data provided by Vicki Lam
EEZIDs_List <- read_excel(paste(Data_Path,"/Spatial/Updated_EEZList_17June2016.xlsx",sep=""))
EEZ_CellID <- read_excel(paste(Data_Path,"/Spatial/EEZ_CellID.xlsx",sep=""))
colnames(EEZ_CellID) <- c("EEZID","INDEX")

# DBEM Coordinate system
Coor <- fread(paste(Data_Path,"/Spatial/Lon_Lat_DBEM.txt",sep=""),header = FALSE)
colnames(Coor) <- c("INDEX","Longitude","Latitude")

# From (/Users/jpalacios/Documents/UBC/Oceans_Project/Distribution/By Cell/By_Cell_Analysis.Rmd)
Raw_DBEM_Results <- data.table::fread(paste(Data_Path,"/Distribution/Raw_Data.csv",sep="")) %>% 
  filter(Species != "Anoplopoma fimbria") #Remove

@@

### North American transboundary **RB stocks** are shifting poleward

@@

# Path to save the plots
Path <- "./Results/North_America"
RCP = c("GFDL26F1","IPSL26F1","MPI26F1") #Low RCP

#### New Future Method to account for mean +- sd of models #
MCP_Data <- DBEM_Corrected %>% 
  mutate(RCP = ifelse(Model %in% RCP,"Low_Emission","High_Emission")) %>% 
  group_by(INDEX,
           Model,
           RCP) %>% 
  summarise_at(vars(Beg_In:End_End), sum,na.rm=T) %>% # Sum the MCP of all species in each index by model
  gather("Year","MCP",Beg_In:End_End) %>% 
  mutate(  # Indicate the three period times
    Period = ifelse(Year >= Beg_In & Year <= Beg_End,"Begining",
                    ifelse(Year >= Mid_In & Year <= Mid_End,"Mid_Century",
                           ifelse(Year >= End_In & Year <= End_End,"The_End_Century",
                                  "Other_Years"))
    )
  ) %>% 
  group_by(INDEX,Model,RCP,Period) %>% 
  summarise(Mean_T_MCP =mean(MCP,na.rm=T)) %>% #Average pero Time period
  filter(Period != "Other_Years") %>% 
  spread(Period,Mean_T_MCP) %>% #Spread to estimate MCP_Change
  mutate_at(vars(Mid_Century, # Devide future values by present
                 The_End_Century),
            funs(ifelse(Begining > 0, ./Begining,.)) #if no species today value is the future
  ) %>% 
  mutate_at(vars(Mid_Century, # Estimate percentage change
                 The_End_Century),
            funs(ifelse(Begining == 0 & . == 0, 0, # If for no change than 0
                        ifelse(Begining == 0 & . > 0, 100, # If changes from 0 to anything then 100
                               round(-(1-.)*100) # Estimate percentage change
                        )
            )
            )
  ) %>% 
  gather("Period","MCP_Change",Mid_Century,The_End_Century) %>% 
  mutate(Direction = ifelse(MCP_Change >= 0, "Positive","Negative")) %>% 
  group_by(INDEX,Period,RCP) %>% # Average results per ESMs
  summarise_at(vars(MCP_Change),
               funs("Mean_ESM_MCP_Chng"=mean,
                    "sd_ESM_MCP_Chng"=sd,
                    "N_ESMs"=n(),
                    "Model_Agreement" = paste(length(unique(Direction)),collapse = ";")
               )
  ) %>%
  mutate(
    Model_Agreement = ifelse(Model_Agreement == 1, "Agree","Disagree"),
    Model_Agreement_sd = ifelse(abs(Mean_ESM_MCP_Chng) < sd_ESM_MCP_Chng*2,"Disagree","Agree")
  ) %>% 
  left_join(DBEM_Coor,
            by="INDEX") %>% 
  mutate_at(vars(Mean_ESM_MCP_Chng,
                 sd_ESM_MCP_Chng),
            funs(ifelse(. > 100, 100,.)
            )  # set everything over 100 to 100
  )

@@

![Fig 2. Percentage change of MCP of transboundary fisheries of North America for mid century (2046-2055) **RA referenced** to 2005-2014 under a A) low emission scenario and; B) high emission scenario. Points represent areas where at least one model do not agree in the direction of change](./Results/Fig2.png)

@@

In consequence, **RA each** country's share-ratio of different species would change in the direction as expected from the effects of ocean warming, regardless of the RCP scenario. Overall, shifts will be amplified under the high emission scenario relative to the low greenhouse emission scenario pathway (Table 2).

Table 2. Average (standard deviation) of 33 transboundary species stock-share of all models for each EEZ and emission scenario for mid and end of the 21st century. Averages might not add to 100% due to model differences (e.g., standard deviation)

@@

# OLD ANALYSIS

@@

# OLD TEXT
